Nonparametric Goodness-of-Fit Tests for Discrete Null Distributions

### 2011-08-25 rjournal/AE.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/ks-test@68 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The revised \code{ks.test()} function supports one-sample tests for discrete
null distributions by allowing the second argument, \code{y}, to be
an empirical cumulative distribution function (an R function
with class \code{ecdf}) or an object of class \code{stepfun} specifying
a discrete distribution.  As in the original version of \code{ks.test()},
the presence of ties in the data (the first argument, \code{x}) generates a
warning unless \code{y} describes a discrete distribution.  
If the sample size is less than or equal to 30, exact $p$-values are provided;
otherwise, asymptotic distributions
are used which are reliable in such cases.
When \code{exact = FALSE} the asymptotic distribution is used and known
to be conservative though imprecise for small samples (see \cite{Conover1972}
for details).

@@

NOTE: Add an example or two with simulated p-values.  Remember to add
stats:: everywhere appropriate for clarity, and ks.test::, etc...

@@

data:  c(0, 1) 
D = 0.5, $p$-value = 0.5
alternative hypothesis: two-sided 
\end{Soutput}
\end{Schunk}
Instead, the value of $D$ given in equation (1)
should be 0 and the associated $p$-value should be 1.  
Our revision of \code{ks.test()}
fixes this problem when the user provides a discrete distribution:
\begin{Schunk} 
\begin{Sinput}
> library(ks.test)
> ks.test(c(0, 1), ecdf(c(0, 1)))
\end{Sinput}
\begin{Soutput}
	One-sample Kolmogorov-Smirnov test

@@

Next, we simulate a sample of size 25 from the discrete uniform distribution on
the integers $\{1, 2, \ldots, 10\}$ and show usage of the
new \code{ks.test()} implementation.  The first is the default two-sided test,
where the exact $p$-value is obtained using the methods of \citet{gleser85}.
\begin{Schunk}
\begin{Sinput}
> set.seed(1)
> x <- sample(1:10, 25, replace = TRUE)
> x
\end{Sinput}
\begin{Soutput}
 [1]  3  4  6 10  3  9 10  7  7  1  3  2  7
[14]  4  8  5  8 10  4  8 10  3  7  2  3
\end{Soutput}
\begin{Sinput}
> ks.test(x, ecdf(1:10))
\end{Sinput}
\begin{Soutput}
	One-sample Kolmogorov-Smirnov test

data:  x 
D = 0.08, $p$-value = 0.9354
alternative hypothesis: two-sided 
\end{Soutput}
\end{Schunk}
Next, we conduct the default one-sided test, where Conover's method
provides the exact $p$-value (up to the numerical precision of the
implementation):
\begin{Schunk}
\begin{Sinput}
> ks.test(x, ecdf(1:10), alternative = "g")
\end{Sinput}
\begin{Soutput}
	One-sample Kolmogorov-Smirnov test

data:  x 
D^+ = 0.04, $p$-value = 0.7731
alternative hypothesis:
the CDF of x lies above the null hypothesis 
\end{Soutput}
\end{Schunk}
In contrast, the option \code{exact=FALSE} results in the
$p$-value obtained by applying the classical Kolmogorov-Smirnov
test, resulting in a conservative $p$-value:
\begin{Schunk}
\begin{Sinput}
> ks.test(x, ecdf(1:10), alternative = "g", 
+         exact = FALSE)
\end{Sinput}
\begin{Soutput}
	One-sample Kolmogorov-Smirnov test

data:  x 
D^+ = 0.04, $p$-value = 0.9231
alternative hypothesis:
the CDF of x lies above the null hypothesis 
\end{Soutput}
\end{Schunk}

A different toy example shows the dangers of using R's existing
\code{ks.test()} function with discrete data:
\begin{Schunk}
\begin{Sinput}
> ks.test(rep(1, 3), ecdf(1:3))
\end{Sinput}
\begin{Soutput}
	One-sample Kolmogorov-Smirnov test

@@

This paper presents the implementation of several nonparametric
goodness-of-fit tests for discrete null distributions.  
In some cases the $p$-values are known to be exact.  In others,
conservativeness in special cases with small $p$-values
has been established.  As a result,
no simulations are necessary for these methods, which were generally
developed during an era when extensive simulations may have been
prohibitively expensive or time-consuming.
However, this does raise the possibility that alternative tests relying upon
modern computational abilities could provide even greater power in certain
situations, a possible avenue for future work.

@@

We are grateful to Le-Minh Ho for his implementation of the algorithm
described in \citet{nieder81} for calculating rectangular probabilities
of uniform order statistics.

>>>>

The revised \code{ks.test()} function supports one-sample tests for discrete
null distributions by allowing the second argument, \code{y}, to be
an empirical cumulative distribution function (an R function
with class \code{ecdf}) or an object of class \code{stepfun} specifying
a discrete distribution.  As in the original version of \code{ks.test()},
the presence of ties in the data (the first argument, \code{x}) generates a
warning unless \code{y} describes a discrete distribution.  
If the sample size is less than or equal to 30, or when \code{exact=TRUE},
exact $p$-values are provided (a warning is issued when the sample size is
greater than 30 due to possible numerical instabilities).
When \code{exact = FALSE} (or when \code{exact} is unspecified and the sample
size is greater than 30) the classical Kolmogorov-Smirnov null distribution
of the test statistic is used and resulting $p$-values are known
to be conservative though imprecise (see \cite{Conover1972}
for details).  In such cases, simulated $p$-values may be desired, produced
by the \code{simulate.p.value=TRUE} option.

@@

@@

data:  c(0, 1) 
D = 0.5, $p$-value = 0.5
alternative hypothesis: two-sided 
\end{Soutput}
\end{Schunk}
Instead, the value of $D$ given in equation (1)
should be 0 and the associated $p$-value should be 1.  
Our revision of \code{ks.test()}
fixes this problem when the user provides a discrete distribution:
\begin{Schunk} 
\begin{Sinput}
> library(ks.test)
> ks.test::ks.test(c(0, 1), ecdf(c(0, 1)))
\end{Sinput}
\begin{Soutput}
	One-sample Kolmogorov-Smirnov test

@@

Next, we simulate a sample of size 25 from the discrete uniform distribution on
the integers $\{1, 2, \ldots, 10\}$ and show usage of the
new \code{ks.test()} implementation.  The first is the default two-sided test,
where the exact $p$-value is obtained using the methods of \citet{gleser85}.
\begin{Schunk}
\begin{Sinput}
> set.seed(1)
> x <- sample(1:10, 25, replace = TRUE)
> x
\end{Sinput}
\begin{Soutput}
 [1]  3  4  6 10  3  9 10  7  7  1  3  2  7
[14]  4  8  5  8 10  4  8 10  3  7  2  3
\end{Soutput}
\begin{Sinput}
> ks.test::ks.test(x, ecdf(1:10))
\end{Sinput}
\begin{Soutput}
	One-sample Kolmogorov-Smirnov test

data:  x 
D = 0.08, $p$-value = 0.9354
alternative hypothesis: two-sided 
\end{Soutput}
\end{Schunk}
Next, we conduct the default one-sided test, where Conover's method
provides the exact $p$-value (up to the numerical precision of the
implementation):
\begin{Schunk}
\begin{Sinput}
> ks.test::ks.test(x, ecdf(1:10),
+   alternative = "g")
\end{Sinput}
\begin{Soutput}
	One-sample Kolmogorov-Smirnov test

data:  x 
D^+ = 0.04, $p$-value = 0.7731
alternative hypothesis:
the CDF of x lies above the null hypothesis 
\end{Soutput}
\end{Schunk}
In contrast, the option \code{exact=FALSE} results in the
$p$-value obtained by applying the classical Kolmogorov-Smirnov
test, resulting in a conservative $p$-value:
\begin{Schunk}
\begin{Sinput}
> ks.test::ks.test(x, ecdf(1:10),
+   alternative = "g", exact = FALSE)
\end{Sinput}
\begin{Soutput}
	One-sample Kolmogorov-Smirnov test

data:  x 
D^+ = 0.04, $p$-value = 0.9231
alternative hypothesis:
the CDF of x lies above the null hypothesis 
\end{Soutput}
\end{Schunk}
The $p$-value may also be estimated via a Monte Carlo simulation:
\begin{Schunk}
\begin{Sinput}
> ks.test::ks.test(x, ecdf(1:10),
+   alternative = "g",
+   simulate.p.value=TRUE, B=10000)
\end{Sinput}
\begin{Soutput}
	One-sample Kolmogorov-Smirnov test

data:  x 
D^+ = 0.04, p-value = 0.7717
alternative hypothesis: 
the CDF of x lies above the null hypothesis 
\end{Soutput}
\end{Schunk}

A different toy example shows the dangers of using R's existing
\code{ks.test()} function with discrete data:
\begin{Schunk}
\begin{Sinput}
> ks.test::ks.test(rep(1, 3), ecdf(1:3))
\end{Sinput}
\begin{Soutput}
	One-sample Kolmogorov-Smirnov test

@@

This paper presents the implementation of several nonparametric
goodness-of-fit tests for discrete null distributions.  
In some cases the $p$-values are known to be exact.  In others,
conservativeness in special cases with small $p$-values
has been established.  Although we provide for Monte Carlo simulated
$p$-values with the new \code{ks.test()},
no simulations may be necessary necessary for these methods; they
were generally
developed during an era when extensive simulations may have been
prohibitively expensive or time-consuming.
However, this does raise the possibility that alternative tests relying upon
modern computational abilities could provide even greater power in certain
situations, a possible avenue for future work.

@@

We are grateful to Le-Minh Ho for his implementation of the algorithm
described in \citet{nieder81} for calculating rectangular probabilities
of uniform order statistics.  We also appreciate the feedback from two
reviewers of this paper as well as from a reviewer of 
\cite{emersonarnold2010}.
